---
title: "Code for 'Bayesian Joint Modelling of Recurrence and Survival: a Conditional Approach'"
author: "Willem van den Boom"
date: "May 14, 2020"
output: html_document
---


In addition to the following R packages, the usage of `rjags` in the comparison with [Paulon et al. (2018)](https://doi.org/10.1093/biostatistics/kxy026) requires [JAGS](https://sourceforge.net/projects/mcmc-jags/) to be installed.

```{r}
for (tmp in c("frailtypack", "matrixStats", "openxlsx", "rjags")) {
  
  if(!tmp %in% rownames(installed.packages())) {
    install.packages(tmp, dependencies = T)
  }
  
}
```


## Read in the data and run the Gibbs sampler

```{r eval=FALSE}
AF_data <- openxlsx::read.xlsx("https://doi.org/10.1371/journal.pone.0217983.s001")


# Recover the actual age from the demeaned version.
AF_data$age <- AF_data$ageM + 64 - max(
  AF_data$ageM[AF_data$ageter == "(54,64]"]
)

# Create binary indicators
AF_data$gender <- AF_data$gender == "female"
AF_data$AFtype <- AF_data$AFtype == "paroxysmal"
AF_data$hypertension <- AF_data$hypertension == "yes"

# Indicator for heart disease
AF_data$heart_disease <- AF_data$heart.fail == "yes" |
  AF_data$heart.valv.dis == "yes" | AF_data$isch.heart.dis == "yes"


# Extract the recurrent events
recurrent_events <- subset(
  x = AF_data,
  subset = !(AF_data$event.type %in% c("endpoint", "Death")),
  select = c(Patient.ID, end)
)

# Only consider patients with at least 2 recurrent events
tmp <- table(recurrent_events$Patient.ID)
IDs <- as.numeric(names(tmp[tmp > 1]))
AF_data <- subset(AF_data, subset = AF_data$Patient.ID %in% IDs)

L <- length(IDs)

# Observed recurrent events
T_obs <- matrix(NA_integer_, nrow = L, ncol = max(tmp) - 1)
n <- rep(NA_integer_, L)

# Survival and censoring times
# The variable is called `c_i` instead of `c`
# as the latter already has a different meaning in R.
c_i <- aggregate(end ~ Patient.ID, data = AF_data, FUN = max)$end

for (i in 1:L) {
  tmp <- recurrent_events$end[recurrent_events$Patient.ID == IDs[i]]
  n[i] <- length(tmp) - 1L
  T_obs[i, 1:n[i]] <- tmp[-1] - tmp[1]
  c_i[i] <- c_i[i] - tmp[1]
}


x <- AF_data[match(IDs, AF_data$Patient.ID), c(
  "age",
  "gender",
  "AFtype",
  "hypertension",
  "heart_disease",
  "Antiarrhythmic"
)]

x$age <- scale(x$age)
x <- as.matrix(x)


# Record whether someone is censored.
cens <- aggregate(event.type ~ Patient.ID, data = AF_data, FUN = function(x) any(x == "endpoint"))$event.type


# We no longer need `IDs`.
rm(IDs, tmp, i)

source("Gibbs_sampler.R")

# Set seed for reproducibility.
set.seed(1)

result <- Gibbs_sampler(
  T_obs = T_obs,
  n = n,
  c_i = c_i,
  cens = cens,
  x = x,
  n_iter = 2e5L,
  burnin = 2e4L,
  thin = 1e1L
)
```

```{r include=FALSE, eval=FALSE}
dir.create("Data", showWarnings = FALSE)
save.image(file = "Data/AF_data.RData", compress = TRUE)
```

```{r include=FALSE, eval=TRUE}
load(file = "Data/AF_data.RData")
```


## Credible intervals

Compute and plot 95% posterior credible intervals for the regression coefficients 
β (solid lines) and γ (dashed lines) of each covariate. 

```{r}
# The number of covariates
q <- ncol(x)

beta_CI <- matrix(data = NA_real_, nrow = q, ncol = 3)
gamma_CI <- matrix(data = NA_real_, nrow = q, ncol = 3)

beta_CI[, 1] <- colMeans(result$beta)
gamma_CI[, 1] <- colMeans(result$gamma)

for (k in 1:q) {
  beta_CI[k, 2:3] <- quantile(result$beta[, k], prob = c(0.025, 0.975))
  gamma_CI[k, 2:3] <- quantile(result$gamma[, k], prob = c(0.025, 0.975))
}


# Make the covariate names prettier for plotting.
colnames(x) <- c(
  "Age",
  "Female",
  "Paroxysmal AF",
  "Hypertension",
  "Heart disease",
  "Anti-arrhythmic med."
)

x_offset <- .15

# Function that draws the credible interval
plot_CI <- function(x, CI, lty = 1) {
  
  # Draw the credible interval
  lines(x = rep(x, 2), y = CI[-1], lty = lty, lwd = 2)
  
  # Add horizontal delimters to both end of the credible interval
  for (y in CI[-1]) lines(
    x = c(x - x_offset / 2, x + x_offset / 2),
    y = rep(y, 2),
    lwd = 2
  )
  
  # Add a dot at the posterior mean
  points(x = x, y = CI[1], pch = 19)
  
}

ylim = range(beta_CI, gamma_CI)


dir.create("Figures", showWarnings = FALSE)
pdf(file = "Figures/plot_CI.pdf", width = 6.5, height = 4)

# Set margins to make room for x axis labels
par(mar = c(8.5, 4, 0, 0) + 0.1)

# Set up empty plot
plot(
  x = 0,
  type = 'n',
  xlim = c(1 - 2 * x_offset, q + 2 * x_offset),
  ylim = ylim,
  xlab = "",
  ylab = "Regression coefficient",
  xaxt = "n"
)

# Add x-axis tick mark text
text(
  x = 1:q,
  y = ylim[1] - 2,
  srt = 90,
  adj = 1,
  labels = colnames(x),
  xpd = TRUE
)

# Add a line at zero
abline(h = 0)

# Plot the q credible intervals
for (k in 1:q) {
  plot_CI(x = k - x_offset, CI = beta_CI[k, ])
  plot_CI(x = k + x_offset, CI = gamma_CI[k, ], lty = 2)
}

dev.off()
```


## Number of events

Compute and plot the CIs.

```{r}
N_CI <- matrix(data = NA_real_, nrow = L, ncol = 3)

N_CI[, 1] <- colMeans(result$N)

for (i in 1:L) N_CI[i, 2:3] <- quantile(result$N[, i], prob = c(0.025, 0.975))


y <- rank(N_CI[,1], ties.method = "random")

y_offset <- 0.35

pdf(file = "Figures/N_CI.pdf", width = 4, height = 7)
par(mar = c(4, 0, 0, 0) + 0.1)

plot(
  x = N_CI[,1],
  y = y,
  pch = ifelse(cens, 19, 1),
  xlim = range(N_CI),
  xlab = expression(italic("N"["i"])),
  ylab = "",
  yaxt = "n"
)

for (i in which(cens)) {
  
  lines(x = N_CI[i,-1], y = rep(y[i], 2), lwd = 2)
  
  # Add an 'x' at the number of observed recurrent events.
  points(x = n[i], y = y[i], pch = 4)
  
}

dev.off()
```


## Posterior predictive distributions of δ, mi0 and mi1

```{r}
n_rep <- 3L
n_recorded <- length(result$M)

m_new <- matrix(NA_real_, nrow = n_rep * n_recorded, ncol = 2)
delta_new <- rep(NA_real_, n_rep * n_recorded)

sigma2_m <- 100
sigma2_delta <- 100

set.seed(1)

for (iter in 1:n_recorded) for (r in 1:n_rep) {
  
  ind <- n_rep * (iter - 1L) + r
  
  if (runif(1) < L / (result$M[iter] + L)) {
    i <- sample.int(L, size = 1)
    m_new[ind, ] <- result$m[iter, i, ]
    delta_new[ind] <- result$delta[iter, i]
  } else {
    m_new[ind, ] <- rnorm(2, sd = sqrt(sigma2_m))
    delta_new[ind] <- rnorm(1, sd = sqrt(sigma2_delta))
  }
  
}


delta_lim <- c(-50, 40)

plot_2D_density <- function(y, ylim, zlim, ylab) {
  
  tmp <- MASS::kde2d(
    x = delta_new,
    y = y,
    h = 2 * c(MASS::bandwidth.nrd(delta_new), MASS::bandwidth.nrd(y)),
    n = 200L,
    lims = c(
      delta_lim + c(-.05, .05) * diff(delta_lim),
      ylim + c(-.05, .05) * diff(ylim)
    )
  )
  
  contour(
    x = tmp$x,
    y = tmp$y,
    z = log(tmp$z),
    xlim = delta_lim,
    ylim = ylim,
    zlim = zlim,
    drawlabels = FALSE,
    xlab = expression(delta[i]),
    ylab = ylab,
  )
  
}

pdf(file = "Figures/m_delta_new.pdf", width = 6.5, height = 3)
par(mfrow = c(1, 2))

plot_2D_density(
  y = m_new[, 1],
  ylim = c(2, 9),
  zlim = c(-8, -3),
  ylab = expression('m'[i1])
)

plot_2D_density(
  y = m_new[, 2],
  ylim = c(-1, 1.5),
  zlim = c(-6, 0),
  ylab = expression('m'[i2])
)

dev.off()
```


## Cluster allocation that minimizes Binder's loss function

```{r}
# Split the Gibbs samples to estimate the transition probabilities
# and minimize Binder's loss on different poterior samples.

n_train <- n_recorded %/% 2L
n_test <- n_recorded - n_train

# Posterior incidence probabilities
p_ij <- matrix(0L, nrow = L, ncol = L)

for (iter in 2L * 1:n_train - 1L) {
  p_ij <- p_ij + outer(result$s[iter, ], result$s[iter, ], "==")
}

p_ij <- p_ij / n_train

Binder_g <- -Inf

lower_tri_ind <- which(lower.tri(matrix(nrow = L, ncol = L)))

for (iter in 2L * 1:n_test) {
  
  tmp <- sum(((p_ij - 0.5) * outer(
    X = result$s[iter, ],
    Y = result$s[iter, ],
    FUN = "=="
  ))[lower_tri_ind])
  
  if (tmp > Binder_g) {
    Binder_g <- tmp
    iter_min <- iter
  }
  
}

s_min <- result$s[iter_min, ]
```


## Sample-wide and cluster-specific summary statistics

```{r}
print(table(n))

Y_obs <- log(t(diff(t(cbind(0, T_obs)))))


pdf(file = "Figures/hist_Y_ij.pdf", width = 6.5, height = 3)
hist(Y_obs, xlab = expression(Y[ij] ~ "(log days)"), main = "")
dev.off()

pdf(file = "Figures/hist_c_i.pdf", width = 6.5, height = 3)
par(mfrow = c(1, 2))
breaks = 2:8
ylim = c(0, 35)

hist(
  x = log(c_i[!cens]),
  xlab = expression(log(S[i]) ~ "(log days)"),
  main = "Uncensored",
  breaks = breaks,
  ylim = ylim
)

hist(
  x = log(c_i[cens]),
  xlab = expression(log(c[i]) ~ "(log days)"),
  main = "Censored",
  breaks = breaks,
  ylim = ylim
)

dev.off()


print(table(s_min))

for (cl in 0:2) {
  
  cat(paste("\nCluster", cl, "\n"))
  
  ind <- if (cl == 0) rep(TRUE, L) else s_min == cl
  print(paste("Number of patients:", sum(ind)))
  print(paste("Proportion censored:", mean(cens[ind])))
  
  tmp <- n[ind & !cens]
  print(paste("Average uncensored N_i:", mean(tmp)))
  print(sd(tmp))
  
  tmp <- colMeans(result$N[, ind])
  print(paste("Average posterior mean of N_i:", mean(tmp)))
  print(sd(tmp))
  
  tmp <- Y_obs[ind, ]
  print(paste("Average uncensored Y_ij:", mean(tmp, na.rm = TRUE)))
  print(sd(tmp, na.rm = TRUE))
  
  tmp <- c(Y_obs[ind, ], colMeans(result$Y_unobs[, ind[cens], ], na.rm = TRUE))
  print(paste("Average posteror mean of Y_ij:", mean(tmp, na.rm = TRUE)))
  print(sd(tmp, na.rm = TRUE))
  
  tmp <- log(c_i[ind & !cens])
  print(paste("Average uncensored log(S):", mean(tmp)))
  print(sd(tmp))
  
  tmp <- colMeans(log(result$S[, ind]))
  print(paste("Average posterior mean of log(S):", mean(tmp)))
  print(sd(tmp))
  
  tmp <- AF_data$age[!duplicated(AF_data$Patient.ID)][ind]
  print(paste("Average age:", mean(tmp)))
  print(sd(tmp))
  
  for (var in colnames(x)[-1]) print(
    paste0("Proportion ", var, ": ", mean(x[ind, var]))
  )
  
}


# Kaplan-Meier plots
S_log_post_mean <- colMeans(log(result$S))

pdf(file = "Figures/K-M_plots.pdf", width = 6.5, height = 3)

plot(
  x = 0,
  ylim = 0:1,
  xlim = c(0, max(S_log_post_mean[s_min <= 2])),
  type = "n",
  xlab = "Log time (log days)",
  ylab = "Proportion surviving"
)

for (cl in 1:2) {
  
  ind <- s_min == cl
  tmp <- sort(S_log_post_mean[ind])
  
  lines(
    x = c(0, rep(tmp, each = 2), tmp[length(tmp)]),
    y = rep(c(1, Vectorize(function (x) mean(tmp > x))(tmp)), each = 2),
    type = 'l',
    lty = cl,
    lwd = 2
  )
  
}

dev.off()
```


## Marginal posterior densities for $\sigma^2$, $\eta^2$ and $\lambda$.

```{r}
pdf(file = "Figures/variances_lambda.pdf", width = 6.5, height = 3)

par(mfrow = c(1, 3))

plot(
  x = density(result$sigma2, to = 15),
  main = "",
  xlab = expression(sigma^2),
  lwd = 2
)

plot(
  x = density(result$eta2, to = 1),
  main = "",
  xlab = expression(eta^2),
  lwd = 2
)

plot(
  x = density(result$lambda),
  main = "",
  xlab = expression(lambda),
  lwd = 2
)

dev.off()
```


## Marginal posterior of the number of clusters

```{r}
pdf(file = "Figures/n_cluster.pdf", width = 6.5, height = 3)

tmp <- table(result$K)

barplot(
  height = tmp / sum(tmp),
  xlab = "Number of clusters",
  ylab = "Posterior probability",
  ylim = c(0, .4)
)

dev.off()
```



## Comparison with the Cox proportional hazards model

```{r}
summary(survival::coxph(survival::Surv(
  time = c_i,
  event = !cens
) ~ x + rowMeans(Y_obs, na.rm = TRUE)))
```


## Comparison with `frailtypack`

The function `frailtyPenal` requires that we use `cluster` and `terminal` instead of `frailtypack::cluster` and `frailtypack::terminal` in its `formula` argument.

```{r}
cluster <- frailtypack::cluster
terminal <- frailtypack::terminal
```


We do not include the covariate `hypertension` as otherwise the method fails to converge.

```{r}
frailtypack::frailtyPenal(
  formula = survival::Surv(
    time = end,
    event = !(event.type %in% c("Death", "endpoint"))
  ) ~ cluster(Patient.ID) + ageM + gender + AFtype + heart_disease +
    Antiarrhythmic + terminal(as.numeric(
      event.type %in% c("Death", "PermAF")
    )),
  formula.terminalEvent = ~ ageM + gender + AFtype + heart_disease +
    Antiarrhythmic,
  data = AF_data,
  jointGeneral = TRUE,
  n.knots = 20,
  kappa = c(1e4, 1e4)
)
```


## Comparison with the Bayesian model from [Paulon et al. (2018)](https://doi.org/10.1093/biostatistics/kxy026)

The BUGS code is based on the code in the supplementary data of [Paulon et al. (2018)](https://doi.org/10.1093/biostatistics/kxy026).

```{r eval=FALSE}
T_n <- rep(NA_real_, L)

for (i in 1:L) {
  T_n[i] <- T_obs[i, n[i]]
}

jags_model <- rjags::jags.model(
  file = textConnection("model{
    
    for (i in 1:L) {
      
      # Likelihood for the n_i observed gap times
      mu[i] <- sum(x[i, ] * beta) + alpha[i]
      
      for (j in 1:n[i]) {
        Y[i,j] ~ dnorm(mu[i], prec[i])
      }
      
      # Survival component for the last (always censored) gap time
      cens_y[i] ~ dinterval(y_last[i], censorLimit_y[i])
      y_last[i] ~ dnorm(mu[i], prec[i])
      
      # Likelihood (censored or not) for the survival component of the model
      cens_u[i] ~ dinterval(U[i], censorLimit_u[i])
      lambda[i] <- sum(x[i, ] * gamma) + psi * alpha[i]
      U[i] ~ dnorm(lambda[i], precU)
      
    }
    
    # Prior specification
    for (k in 1:q) {
      beta[k] ~ dnorm(0,1.0E-2)
      gamma[k] ~ dnorm(0,1.0E-2)
    }
    
    precU ~ dgamma(2.01,1.01)
    psi ~ dnorm(0,1.0E-2)
    M ~ dunif(0.3, 5.0)
    
    # Stick-breaking construction
    p[1] <- v[1]
    
    for (h in 2:Napp) {
      p[h] <- v[h] * (1 - v[h - 1]) * p[h - 1] / v[h - 1]
    }
    
    p.sum <- sum(p[])
    
    # Sample from the DP prior for alpha and prec
    for (h in 1:Napp) {
      
      theta_1[h] ~ dnorm(0,1.0E-2) # locations of alpha
      theta_2[h] ~ dgamma(2.01,1.01) # locations of prec
      v[h] ~ dbeta(1, M)
      
      # Scaling to ensure sum to 1
      pp[h] <- p[h]/p.sum
      
    }
    
    for (i in 1:L) {
      S[i] ~ dcat(pp[]) # latent variable: cluster indicator
      alpha[i] <- theta_1[S[i]]
      prec[i] <- theta_2[S[i]]
    }
    
    # Predictive distributions
    S_new ~ dcat(pp[])
    alpha_new <- theta_1[S_new]
    prec_new <- theta_2[S_new]
  
  }"),
  data = list(
    Y = Y_obs,
    U = ifelse(cens, NA, log(c_i)),
    x = x,
    Napp = 40L,
    n = n,
    q = q,
    L = L,
    cens_u = cens,
    censorLimit_u = log(c_i),
    cens_y = rep(1L, L),
    censorLimit_y = log(c_i - T_n)
  ),
  # Set seed for reproducibility.
  inits = list(.RNG.name = "base::Wichmann-Hill", .RNG.seed = 1)
)

rjags:::update.jags(object = jags_model, n.iter = 1e4)

result_Paulon <- rjags::jags.samples(
  model = jags_model,
  variable.names = c(
    "beta",
    "gamma",
    "psi",
    "precU",
    "alpha",
    "alpha_new",
    "prec_new"
  ),
  n.iter = 6e4,
  thin = 12
)
```

```{r include=FALSE, eval=FALSE}
save(result_Paulon, file = "Data/result_Paulon.RData", compress = TRUE)
```

```{r include=FALSE, eval=TRUE}
load(file = "Data/result_Paulon.RData")
```


Plot the marginal posterior on the number of clusters.

```{r}
pdf(file = "Figures/n_cluster_Paulon.pdf", width = 6.5, height = 3)

tmp <- table(
  apply(result_Paulon$alpha, MARGIN = 2, FUN = function (x) length(unique(x)))
)

barplot(
  height = tmp / sum(tmp),
  xlab = "Number of clusters",
  ylab = "Posterior probability",
  ylim = c(0, .4)
)

dev.off()
```


Plot the credible intervals for the regression coefficients.

```{r}
beta_CI_Paulon <- matrix(data = NA_real_, nrow = q, ncol = 3)
gamma_CI_Paulon <- matrix(data = NA_real_, nrow = q, ncol = 3)

beta_CI_Paulon[, 1] <- rowMeans(result_Paulon$beta)
gamma_CI_Paulon[, 1] <- rowMeans(result_Paulon$gamma)

for (k in 1:q) {
  
  beta_CI_Paulon[k, 2:3] <- quantile(
    x = result_Paulon$beta[k, , 1],
    prob = c(0.025, 0.975)
  )
  
  gamma_CI_Paulon[k, 2:3] <- quantile(
    x = result_Paulon$gamma[k, , 1],
    prob = c(0.025, 0.975)
  )
  
}


ylim = range(beta_CI_Paulon, gamma_CI_Paulon)

pdf(file = "Figures/plot_CI_Paulon.pdf", width = 6.5, height = 4)

# Set margins to make room for x axis labels.
par(mar = c(8.5, 4, 0, 0) + 0.1)

# Set up empty plot
plot(
  x = 0,
  type = 'n',
  xlim = c(1 - 2 * x_offset, q + 2 * x_offset),
  ylim = ylim,
  xlab = "",
  ylab = "Regression coefficient",
  xaxt = "n"
)

# Add x-axis tick mark text.
text(
  x = 1:q,
  y = ylim[1] - .75,
  srt = 90,
  adj = 1,
  labels = colnames(x),
  xpd = TRUE
)

# Add a line at zero.
abline(h = 0)

# Plot the q credible intervals for both sets of regression coefficients.
for (k in 1:q) {
  plot_CI(x = k-x_offset, CI = beta_CI_Paulon[k, ])
  plot_CI(x = k+x_offset, CI = gamma_CI_Paulon[k, ], lty = 2)
}

dev.off()
```


Plot the marginal posterior of $\psi$.

```{r}
pdf(file = "Figures/psi_Paulon.pdf", width = 6.5, height = 3)

plot(
  x = density(result_Paulon$psi),
  main = "",
  xlab = expression(psi),
  lwd = 2
)

dev.off()
```