---
title: "Simulation"
author: "Willem van den Boom"
date: "June 21, 2020"
output: html_document
---


## Generate data according to the model

The following uses parallel computing which I have only tested on macOS and Linux and might therefore not work on Windows operating systems.

```{r eval=FALSE}
# For parallel computing
if(!'doParallel' %in% rownames(installed.packages())) {
  install.packages('doParallel', dependencies = T)
}

censoring_rate <- c(0, 0.5, 0.8, 0.9)
n_cr <- length(censoring_rate)


# Set seed for reproducibility.
set.seed(1)

L <- 150L # Number of individuals
s <- rep(1:3, 50) # Cluster allocation
K <- max(s) # Number of clusters

# Specify the random effects.
m <- matrix(data = NA_real_, nrow = L, ncol = 2)
m[, 1] <- s
for (h in 1:K) m[s == h, 2] <- (h - 2) * 0.8
delta <- s + 4

sigma2 <- 1
eta2 <- 1
lambda <- 7

N <- as.integer(qpois(
  p = runif(n = L, min = ppois(q = 0, lambda = lambda)),
  lambda = lambda
))
N_max <- max(N)

# The exponentiation can lead to `Inf`.
# We therefore use `pmin`.
S <- pmin(exp(rnorm(
  n = L,
  mean = delta,
  sd = sqrt(eta2)
)), 1e100)


# Generate Y from its distribution conditional on S using rejection sampling.
Y <- matrix(data = NA_real_, nrow = L, ncol = N_max)
T_all <- Y
for (i in 1:L) T_all[i, N[i]] <- Inf

for (i in 1:L) while(T_all[i, N[i]] > S[i]) for (j in 1:N[i]) {
  
  Y[i, j] <- rnorm(
    n = 1,
    mean = m[i, 1] + if (j > 1) m[i, 2] * (Y[i, j - 1L] - m[i, 1]) else 0,
    sd = sqrt(sigma2)
  )
  
  T_all[i, j] <- exp(Y[i, j]) + if (j > 1) T_all[i, j - 1L] else 0
  
}


# Generate data and run the Gibbs sampler for the r-th censoring rate.
run_Gibbs <- function (r) {
  
  # Set seed for reproducibility.
  set.seed(1)
  
  T_obs <- T_all
  n <- N
  c_i <- S
  
  cens <- rep(FALSE, L)
  cens[sample(L, size = round(censoring_rate[r] * L))] <- TRUE
  
  for (i in which(cens)) {
    c_i[i] <- runif(n = 1, min = T_all[i, 1], max = S[i])
    tmp <- T_all[i, 1:N[i]] <= c_i[i]
    T_obs[i, !tmp] <- NA_real_
    n[i] <- sum(tmp)
  }
  
  T_obs <- T_obs[, 1:max(n)]
  
  source("Gibbs_sampler.R")
  
  result <- Gibbs_sampler(
    T_obs = T_obs,
    n = n,
    c_i = c_i,
    cens = cens,
    n_iter = 2e5L,
    burnin = 2e4L,
    thin = 1e1L
  )
  
  return(list(result = result, n = n, cens = cens))
  
}


# Setup for parallel computing
# The `setup_strategy` argument is set as a workaround to a bug in RStudio
# on macOS with R 4.0:
# https://github.com/rstudio/rstudio/issues/6692#issuecomment-619645114
cl <- parallel::makeCluster(n_cr, setup_strategy = "sequential")
doParallel::registerDoParallel(cl)

# Fit the Gibbs sampler for each of the simulated data sets.
fit_list <- foreach::`%dopar%`(foreach::foreach(r = 1:n_cr), run_Gibbs(r))
parallel::stopCluster(cl)

result_c <- list()
n_c <- list()
cens_c <- list()

for (r in 1:n_cr) {
  result_c[[r]] <- fit_list[[r]]$result
  n_c[[r]] <- fit_list[[r]]$n
  cens_c[[r]] <- fit_list[[r]]$cens
}

rm(fit_list)
```


```{r include=FALSE, eval=FALSE}
dir.create("Data", showWarnings = FALSE)
save.image(file = "Data/simulation.RData", compress = TRUE)
```

```{r include=FALSE, eval=TRUE}
load(file = "Data/simulation.RData")
```





## Number of events

Compute and plot the CIs.

```{r}
titles <- paste0(100 * censoring_rate, "% censored")

pdf(
  file = paste0("Figures/N_CI_simulation.pdf"),
  width = 10,
  height = 15
)

par(mfrow = c(1, n_cr - 1), mar = c(4, 0, 3, 0))

y <- rank(
  x = 10000L * N + 100L * n_c[[4]] + colMeans(result_c[[4]]$N),
  ties.method = "random"
)

for (r in 2:n_cr) {
  
  result <- result_c[[r]]
  n <- n_c[[r]]
  cens <- cens_c[[r]]

  N_CI <- matrix(data = NA_real_, nrow = L, ncol = 3)
  
  N_CI[, 1] <- colMeans(result$N)
  
  for (i in 1:L) N_CI[i, 2:3] <- quantile(result$N[, i], prob = c(0.025, 0.975))
  
  # Plot an 'o' at the actual number of events.
  plot(
    x = N,
    y = y,
    pch = 1,
    xlim = c(1, 17),
    xlab = expression(italic("N"["i"])),
    ylab = "",
    yaxt = "n",
    main = titles[r]
  )
   
  for (i in which(cens)) {
    
    lines(x = N_CI[i,-1], y = rep(y[i], 2), lwd = 2)
    
    # Add an 'x' at the number of observed recurrent events.
    points(x = n[i], y = y[i], pch = 4)
    
    # Add a solid dot at the posterior mean.
    points(x = N_CI[i, 1], y = y[i], pch = 19)
    
  }

}

dev.off()
```


## Posterior predictive distributions of δ, mi0 and mi1

```{r}
pdf(file = "Figures/m_delta_new_simulation.pdf", width = 6.5, height = 9)
par(mfrow = c(n_cr, 2))

for (rr in 1:n_cr) {
  
  result <- result_c[[rr]]
  
  n_rep <- 3L
  n_recorded <- length(result$M)
  
  m_new <- matrix(NA_real_, nrow = n_rep * n_recorded, ncol = 2)
  delta_new <- rep(NA_real_, n_rep * n_recorded)
  
  sigma2_m <- 100
  sigma2_delta <- 100
  
  set.seed(1)
  
  for (iter in 1:n_recorded) for (r in 1:n_rep) {
    
    ind <- n_rep * (iter - 1L) + r
    
    if (runif(1) < L / (result$M[iter] + L)) {
      i <- sample.int(L, size = 1)
      m_new[ind, ] <- result$m[iter, i, ]
      delta_new[ind] <- result$delta[iter, i]
    } else {
      m_new[ind, ] <- rnorm(2, sd = sqrt(sigma2_m))
      delta_new[ind] <- rnorm(1, sd = sqrt(sigma2_delta))
    }
    
  }
  
  
  delta_lim <- c(2.5, 14)
  
  plot_2D_density <- function(y, ylim, zlim, ylab) {
    
    tmp <- MASS::kde2d(
      x = delta_new,
      y = y,
      h = 2 * c(MASS::bandwidth.nrd(delta_new), MASS::bandwidth.nrd(y)),
      n = 200L,
      lims = c(
        delta_lim + c(-.05, .05) * diff(delta_lim),
        ylim + c(-.05, .05) * diff(ylim)
      )
    )
    
    contour(
      x = tmp$x,
      y = tmp$y,
      z = log(tmp$z),
      xlim = delta_lim,
      ylim = ylim,
      zlim = zlim,
      drawlabels = FALSE,
      xlab = expression(delta[i]),
      ylab = ylab,
      main = titles[rr]
    )
    
  }
  
  
  plot_2D_density(
    y = m_new[, 1],
    ylim = c(-1, 5),
    zlim = c(-6, -1),
    ylab = expression('m'[i1])
  )
  
  plot_2D_density(
    y = m_new[, 2],
    ylim = c(-2, 2.5),
    zlim = c(-6, -1),
    ylab = expression('m'[i2])
  )
  
}

dev.off()
```


## Marginal posterior densities for $\sigma^2$, $\eta^2$ and $\lambda$.

```{r}
pdf(file = "Figures/variances_lambda_simulation.pdf", width = 6.5, height = 9)

par(mfrow = c(n_cr, 3))

for (r in 1:n_cr) {
  
  plot(
    x = density(result_c[[r]]$sigma2, from = 0.9, to = 1.4),
    main = "",
    xlab = expression(sigma^2),
    lwd = 2
  )
  
  plot(
    x = density(result_c[[r]]$eta2, from = 0, to = 6),
    main = titles[r],
    xlab = expression(eta^2),
    lwd = 2
  )
  
  plot(
    x = density(result_c[[r]]$lambda, from = 6.2, to = 9),
    main = "",
    xlab = expression(lambda),
    lwd = 2
  )
  
}

dev.off()
```


## Marginal posterior of the number of clusters

```{r}
pdf(file = "Figures/n_cluster_simulation.pdf", width = 6.5, height = 3)

par(mfrow = c(1, n_cr))

for (r in 1:n_cr) {
  
  tmp <- table(factor(result_c[[r]]$K, levels = 3:7))
  
  barplot(
    height = tmp / sum(tmp),
    xlab = "Number of clusters",
    ylab = "Posterior probability",
    ylim = c(0, 1),
    main = titles[r]
  )

}

dev.off()
```